---
title: "Ambika HDAC inhibitor screen analysis"
author: "F. Heigwer"
date: "8/17/2018"
output:
  html_document: default
  pdf_document: 
    keep_tex: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(FitAR)
library(corrr)
library(ggrastr)
library(ggrepel)
library(broom)
```
#Lets define some themes

This we do as a quality of life step to give all figures that we produce a common look and feel.

## B110 Theme

```{r theme, include=FALSE}

theme_b110 <- function(){
  theme_classic() +
  theme(
    axis.text = element_text(size = 10), 
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12,hjust = 0.5,face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "bottom"
    )
}

```

## B110 Colors

```{r colors}

sgi_blue    = '#5087C8'
sgi_yellow1 = '#F2EE35'
sgi_yellow2 = '#FED98E'
b110_grey   = '#808080'
b110_grey_light   = '#909090'
b110_transparent_black = scales::alpha('#000000',0.5)
google_red = '#dd4b39'
google_green = '#0F9D58'
google_yellow = '#F4B400'
google_blue = '#4285F4'

```

## Data input

Lets check how the data looks like and what the objectives of the analyses are.

We conclude that we have data for two cell lines, in two different passages each. Each was measured for the abundance of 146 different analytes between two treatments. These are however always a little bit different.

```{r data_input}

treat1<-read_delim(file = "../raw_data_counts/counts_B_P_U_screen/P1_S7_L001_R1_001.txt",delim = "\t")
treat2<-read_delim(file = "../raw_data_counts/counts_B_P_U_screen/P2_S8_L001_R1_001.txt",delim = "\t")

treatment<-"P"

system(paste("mkdir",treatment))

dmso1<-read_delim(file = "../raw_data_counts/counts_B_P_U_screen/DMSO1_S1_L001_R1_001.txt",delim = "\t")
dmso2<-read_delim(file = "../raw_data_counts/counts_B_P_U_screen/DMSO2_S2_L001_R1_001.txt",delim = "\t")

mydata<-treat1 %>% 
  left_join(treat2,by="sgRNA") %>% 
  left_join(dmso1,by="sgRNA") %>% 
  left_join(dmso2,by="sgRNA") %>%
  extract(sgRNA,c("sgRNA","gene_id"),"((\\w+)_.*)") 

names(mydata)<-c("sgRNA","gene_id","treat_1","treat_2","dmso_1","dmso_2")

mydata %<>%
  gather(treatment,count,-sgRNA,-gene_id) %>%
  extract(treatment,c("treatment","replicate"),"(\\w+)_(\\d)")

anno<-read_delim("../meta_data/gene_names.txt",delim="\t")

```

## Lets normalize the data

```{r Normalization, echo=FALSE}

mydata %<>% 
  mutate(count=if_else(count>10,count,NA_real_)) %>%    #filter low read counts
  group_by(treatment,replicate) %>% #group by treatment and replicate
  mutate(normval=asinh(count/median(count,na.rm = T)))

```

## Lets do some data QC

```{r QC, echo=FALSE}

#on sgRNA level
p1 <- mydata %>% 
  select(-count) %>%
  spread(replicate,normval) %>%
  ggplot(aes(x=`1`,y=`2`)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~treatment,scales = "free") +
    theme_b110()

ggsave("../graphics/normalized_readcount_replicates_sgRNA_Panobinostat.pdf",width = 8,height = 8)

print(p1)

mydata %>% 
  select(-count) %>%
  spread(replicate,normval) %>%
  split(.$treatment) %>%
  lapply(function(x){x %>% ungroup() %>% select(`1`,`2`) %>% correlate() %>% rearrange() %>% shave() %>% fashion()})

# from this analysis we get a replicate correlation of 0.88 and 0.91 for the DMSO and Panobinostat treated samples respectively

#on gene level
p1 <- mydata %>% 
  select(-count) %>%
  group_by(treatment,gene_id,replicate) %>%
  summarise(mean_normval=mean(normval,na.rm=T)) %>%
  spread(replicate,mean_normval) %>%
  ggplot(aes(x=`1`,y=`2`)) +
    geom_point() +
    geom_smooth(method = "lm")+
    facet_wrap(~treatment,scales = "free") +
    theme_b110()

ggsave("../graphics/normalized_readcount_replicates_genes_Panobinostat.pdf",width = 8,height = 8)

print(p1)

mydata %>% 
  group_by(treatment,gene_id,replicate) %>%
  summarise(mean_normval=mean(normval,na.rm=T)) %>%
  spread(replicate,mean_normval) %>%
  split(.$treatment) %>%
  lapply(function(x){x %>% ungroup() %>% select(`1`,`2`) %>% correlate() %>% rearrange() %>% shave() %>% fashion()})
  
# on gene level we get a replicate read count correlation of 0.94 and 0.97 for DMSO and Panobinostat treated samples respectively
```


## QC looks very nice and good so lets calls some foldchanges and hits thereof

```{r hitcalling, echo=FALSE}

#we average replicates and plot treated versus untreated samples
#sgRNA level
p1 <- mydata %>% 
  group_by(sgRNA,gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ggplot(aes(x=dmso,y=treat)) +
    geom_point() +
    theme_b110()

ggsave("../graphics/readcount_dmso_vs_vorinostat_sgRNA_Panobinostat.pdf",width = 8,height = 8)

print(p1)


#gene level
p1 <- mydata %>% 
  group_by(gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ggplot(aes(x=dmso,y=treat)) +
    geom_point() +
    geom_abline(slope = 1,intercept = c(0,0)) +
    theme_classic()+
    theme_b110()

ggsave("../graphics/readcount_dmso_vs_vorinostat_genes_Panobinostat.pdf",width = 8,height = 8)

print(p1)

#lets call some foldchanges and hits
tmp<-mydata %>% 
  group_by(sgRNA,gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ungroup() %>%
  mutate(logfc=log2(treat/dmso)) 

stats<- tmp %>%
  group_by(gene_id) %>%
  do(
   wilcox.test(.$logfc,tmp$logfc[tmp$gene_id=="random"],alternative = "t") %>% tidy()
  ) %>%
  ungroup() %>%
  mutate(fdr=p.adjust(p.value,method="BH"))

final_sg<- tmp %>% left_join(stats) %>% mutate(significant=if_else(fdr<0.1,"significant","non-significant"))

final_gene<-
  final_sg %>%
  group_by(gene_id) %>%
  summarise(dmso=mean(dmso,na.rm = T),treat=mean(treat,na.rm = T),logfc=mean(logfc,na.rm = T),p.value=unique(p.value),fdr=unique(fdr),significant=unique(significant))

#gene level
p1 <- final_gene %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  ggplot(aes(x=dmso,y=treat,color=significant,label=labels)) +
    geom_point() +
    geom_abline(slope = 1,intercept = c(0,0)) +
    theme_classic() +
    geom_label_repel()+
    scale_color_manual(values = c(b110_grey_light,google_blue))


ggsave("../graphics/readcount_dmso_vs_vorinostat_genes_stats_Panobinostat.pdf",width = 8,height = 8)

print(p1)

#gene level
p1 <- final_gene %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  ggplot(aes(x=logfc,y=-log10(p.value),color=significant,label=labels)) +
    geom_point() +
    theme_classic() +
    geom_label_repel() +
    scale_color_manual(values = c(b110_grey_light,google_blue))

ggsave("../graphics/volcano_stat_colored_Panobinostat.pdf",width = 8,height = 8)

print(p1)

final_gene %>% 
  left_join(anno) %>% 
  write_delim("../results/GeneLevelData_Panobinostat.tsv",delim = "\t")


p1 <- final_sg %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  filter(!is.na(labels)) %>%
  mutate(cols=factor(sign(logfc))) %>%
  ggplot(aes(x=reorder(sgRNA,logfc),y=logfc,fill=cols)) +
    geom_bar(stat = "identity",position = "dodge")  +
  facet_wrap(~gene_symbol,scales = "free") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    scale_fill_manual(values = c(google_blue,google_red))


ggsave("../graphics/EHMT_sgRNAs_colored_Panobinostat.pdf",width = 8,height = 8)

print(p1)

final_sg %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  filter(!is.na(labels)) %>%
  mutate(cols=factor(sign(logfc))) %>% 
  write_delim("../results/EHMT1_EHMT2_sgRNALevelData_Panobinostat.tsv",delim = "\t")

final_sg %>% 
  left_join(anno) %>% 
  filter(gene_id == "random") %>%
  write_delim("../results/random_sgRNALevelData_Panobinostat.tsv",delim = "\t")

p1 <- final_sg %>% 
  left_join(anno) %>% 
  filter(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B") | gene_id %in% c("random")) %>%
  mutate(labels = if_else(is.na(gene_symbol),gene_id,gene_symbol)) %>%
  mutate(cols = factor(sign(logfc))) %>%
  select(labels,sgRNA,logfc,cols) %>%
  drop_na() %>%
  ggplot(aes(x = reorder(sgRNA,logfc), y = logfc, fill = cols)) +
    geom_bar(stat = "identity",position = "dodge")  +
  facet_wrap(~labels,scales = "free_x") +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
    scale_fill_manual(values = c(google_blue,google_red))


ggsave("../graphics/EHMT_random_sgRNAs_colored_Panobinostat.pdf",width = 8,height = 8)

print(p1)
```

# Session info

```{r}
writeLines(capture.output(sessionInfo()), paste("../results/SessionInfo_Panobinostat.txt",sep = ""))
```
