---
title: "Pairwise interactions of Wnt-ligands and FZD receptors"
author: "Oksana Voloshanenko"
date: "5/8/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown script contains code to create a heatmap visualizing pairwise interactions between Wnt ligands and FZD receptors as measured and described in Voloshanenko *et al.*, 2017.

# Loading Dependencies

The script depends on the pheatmap package to draw the heatmap and the RColorBrewer package to generate a color pallette. We first load these packages into the R environment.

```{r, results='hide', warning=F, message=F}
library(pheatmap)
library(RColorBrewer)
```

# Loading the data

We next load the data from a tab separated text file. The first column contains the row names of the table.

```{r, results='hide', warning=F, message=F}
data <- read.delim('WntFz_map.txt', row.names = 1)
```

# Drawing the heatmap

We define a shades of gray color pallette and finally draw the heatmap clustering both rows and columns.

```{r, results='hide', warning=F, message=F}
colors <- colorRampPalette(c("gray100", "gray90", "gray80", "gray70",
                             "gray60", "gray50", "gray40", "gray30",
                             "gray20", "gray10", "gray1"))(200)
pheatmap(as.matrix(data), cluster_rows = TRUE, cluster_cols = TRUE, color=colors)
```

# Session info

```{r}
sessionInfo()
```
