---
title: "Digi-West analysis"
author: "F. Heigwer"
date: "8/17/2018"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, message=TRUE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(packrat)
library(tidyverse)
library(readxl)
library(limma)


theme_b110 <- function(){
  theme_classic() +
  theme(
    axis.text = element_text(size = 16), 
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 22,hjust = 0.5,face = "bold"),
    legend.title = element_text(size = 22),
    legend.text = element_text(size = 16),
    legend.position = "bottom"
    )
}

```

## B110 Colors

```{r colors, message=TRUE, warning=FALSE}
sgi_blue    = '#5087C8'
sgi_yellow1 = '#F2EE35'
sgi_yellow2 = '#FED98E'
b110_grey   = '#808080'
b110_grey_light   = '#909090'
b110_transparent_black = alpha('#000000',0.5)
google_red = '#dd4b39'
google_green = '#0F9D58'
google_yellow = '#F4B400'
google_blue = '#4285F4'

```

## Data input

Lets check how the data looks like and what the objectives of the analyses are.

We conclude that we have data for two cell lines, in two different passages each. Each was measured for the abundance of 146 different analytes between two treatments. These are however always a little bit different.

```{r data_input, message=TRUE, warning=FALSE}
mydata <- read_excel('../raw_data/DigiWest results corrected3 and double checked 191229_adapted.xlsx',col_names = T)

mydata %<>% 
  select(process = Pathway,everything()) %>% 
  gather(treatment,value,-Analyte,-process) %>% 
  extract(treatment,c("cellline","treatment"),"(\\w+ \\w+) (.+)") %>%
  extract(cellline,c("cellline","passage"),"(\\w+) (\\w+)") %>% 
  mutate(treatment = if_else(treatment != 'Combi',"DMSO",treatment)) %>% 
  drop_na() %>%
  distinct() %>%
  spread(treatment,value) %>%
  mutate(fc = log2(Combi/DMSO),process = if_else(grepl("OTHER",process),"OTHER",process))
  
```

## Lets vizualize the data

We vizualize the foldchange (log2 Combi/DMSO) of each analyte per cell line.

```{r overview_plots, echo=FALSE, fig.height=40, fig.width=10, message=TRUE, warning=FALSE}

p1 <- mydata %>% 
  group_by(Analyte,cellline) %>%
  mutate(meanfc = mean(fc)) %>%
  ungroup() %>%
  mutate(Analyte = factor(Analyte,levels = unique(Analyte[order(process)]))) %>%
  ggplot(aes(x = Analyte,y = fc)) +
    geom_hline(yintercept = 0) +
    geom_bar(aes(y = meanfc,fill = process),stat = "identity",position = "dodge") +
    geom_point() +
    coord_flip() +
    facet_wrap(~cellline) +
    theme_classic()

ggsave("../graphics/data_overview.pdf",width = 8,height = 24)

print(p1)

```

# filter by foldchange to be at least 0.4 log2FC

```{r plots, echo=FALSE, message=TRUE, warning=FALSE}

mysplit <- mydata %>% split(.$cellline)

p1 <- mydata %>% 
  group_by(Analyte) %>%
  mutate(meanfc = mean(fc)) %>%
  ungroup() %>%
  mutate(Analyte = factor(Analyte,levels = unique(Analyte[order(process)]))) %>%
  mutate(process = if_else(process %in% c("APOPTOSIS","AUTOPHAGY","CELL CYCLE"),process,"other")) %>%
  filter(abs(meanfc) > .4) %>%
  ggplot(aes(x = reorder(Analyte,-fc,mean),y = fc)) +
    geom_hline(yintercept = 0) +
    geom_bar(aes(y = meanfc,fill = process),stat = "identity",position = "dodge") +
    geom_point() +
    facet_grid(rows = "cellline") +
    theme_classic() +
    scale_fill_manual(values = c("red","blue","yellow","gray")) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("all data log2 foldchange >0.4 bothlines")

ggsave("../graphics/data_overview_cutoff_fc_04.pdf",width = 24,height = 8)

print(p1)

p1 <- mysplit$HCT116 %>% 
  group_by(Analyte) %>%
  mutate(meanfc = mean(fc)) %>%
  ungroup() %>%
  mutate(Analyte = factor(Analyte,levels = unique(Analyte[order(process)]))) %>%
  mutate(process = if_else(process %in% c("APOPTOSIS","AUTOPHAGY","CELL CYCLE"),process,"other")) %>%
  filter(abs(meanfc) > .4) %>%
  filter(!(Analyte %in% c('A-Raf - p- Y301/Y302 # 1844 [68kDa]','Rb  BG/MP # 046 [48kDa]','Cyclin A  TK # 103 [57kDa]','LC3B 1:200   [14kDa]','APG3L  # 0608 [36kDa]'))) %>%
  ggplot(aes(x = reorder(Analyte,-fc,mean),y = fc)) +
    geom_hline(yintercept = 0) +
    geom_bar(aes(y = meanfc,fill = process),stat = "identity",position = "dodge") +
    geom_point() +
    theme_classic() +
    scale_fill_manual(values = c("red","blue","yellow","gray")) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("HCT116 log2 foldchange >0.4")

ggsave("../graphics/HCT116_log2_foldchange.pdf",width = 24,height = 8)

print(p1)

p1 <- mysplit$HT29 %>% 
  group_by(Analyte) %>%
  mutate(meanfc =  mean(fc)) %>%
  ungroup() %>%
  mutate(Analyte =  factor(Analyte,levels = unique(Analyte[order(process)]))) %>%
  mutate(process = if_else(process %in% c("APOPTOSIS","AUTOPHAGY","CELL CYCLE"),process,"other")) %>%
  filter(abs(meanfc) > 0.4) %>%
  filter(!(Analyte %in% c('p38 MAPK - p- T180/Y182 # 1223 [45kDa]','A-Raf - p- Y301/Y302 # 1844 [68kDa]','Rb  BG/MP # 046 [48kDa]','Cyclin A  TK # 103 [57kDa]','LC3B 1:200   [16kDa]','LC3B 1:200   [14kDa]','APG3L  # 0608 [36kDa]'))) %>%
  ggplot(aes(x = reorder(Analyte,-fc,mean),y = fc)) +
    geom_hline(yintercept = 0) +
    geom_bar(aes(y = meanfc,fill = process),stat = "identity",position = "dodge") +
    geom_point() +
    theme_classic() +
    scale_fill_manual(values = c("red","blue","yellow","gray")) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("HT29 log2 foldchange >0.4")

ggsave("../graphics/HT29_log2_foldchange.pdf",width = 24,height = 8)

print(p1)

HCT116_FDR <-
  mysplit$HCT116 %>% 
  select(-Combi,-DMSO) %>% 
  spread(passage,fc) %>% 
  select_if(is.numeric) %>% 
  lmFit() %>% 
  limma::eBayes() %>% 
  .$p.value %>% 
  p.adjust(method = "BH")

HCT116_data <- mysplit$HCT116 %>% select(-Combi,-DMSO) %>% spread(passage,fc) %>% mutate(fdr = HCT116_FDR)

HT29_FDR <- 
  mysplit$HT29 %>% 
  select(-Combi,-DMSO) %>% 
  spread(passage,fc) %>% 
  select_if(is.numeric) %>% 
  lmFit() %>% 
  limma::eBayes() %>% 
  .$p.value %>% 
  p.adjust(method = "BH")

HT29_data <-
  mysplit$HT29 %>% 
  select(-Combi,-DMSO) %>% 
  spread(passage,fc) %>% 
  mutate(fdr = HT29_FDR)

HCT116_data %>%
  arrange(fdr) %>%
  write_delim("../results/HCT116_data_digiwest_28.02.2020.tsv",delim = "\t")

HT29_data %>%
  arrange(fdr) %>%
  write_delim("../results/HT29_data_digiwest_28.02.2020.tsv",delim = "\t")


```

# reproducibility

```{r reproducibility, message=TRUE, warning=FALSE}
p1 <- mydata %>% 
  group_by(Analyte,cellline,passage) %>%
  mutate(meanfc = mean(fc)) %>%
  select(process,Analyte,cellline,passage,meanfc) %>%
  distinct() %>%
  spread(passage,meanfc) %>%
  ggplot(aes(x = R1, y = R2)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~cellline, scales = "free") +
    theme_b110()
 
ggsave("../graphics/reproducibility.pdf",width = 16,height = 8)

print(p1)
```

# Session info

```{r, message=TRUE, warning=FALSE}
writeLines(capture.output(sessionInfo()), paste("../results/SessionInfo.txt",sep = ""))
```
