---
title: "Gene expression heat maps"
author: "Benedikt Rauscher"
date: "9/22/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r, echo=F, message=F, warning=F, results='hide'}
library(openxlsx)
library(pheatmap)
library(tidyverse)
```

# Loading the data

We first load the expression data from Excel files.

```{r, results='hide', warning=F, message=F}
## Table S1
ts1 <- read.xlsx('Table S1.xlsx', rowNames = T, startRow = 4)
## canonical main
canonical <- read.xlsx('canonical main.xlsx', rowNames=T)
## canonical N2
canonical_n2 <- read.xlsx('canonical N2.xlsx', rowNames=T)
## canonical N3
canonical_n3 <- read.xlsx('canonical N3.xlsx', rowNames=T)
```

# Draw heat maps

## Data normalization 

We normalize each dataset by standardizing the per-gene expression values. This highlights differences in gene expression across different samples but loses information about the base expression level of each gene.

```{r, results='hide', warning=F, message=F}
## normalize each data set
data_norm <- lapply(list(ts1, canonical, canonical_n2, canonical_n3), function(d){
  ## save names
  cn <- colnames(d)
  ## z-score normalization
  norm <- t(apply(scale(log(d)), 1, function(x) scale(x)))
  colnames(norm) <- cn
  return(t(norm))
})
```

## Visualization

We can now visualize the normalized expression values as heatmaps using the `pheatmap` package.

```{r, results='hide', warning=F, message=F}
## heatmap colours
cols <- rev(colorRampPalette(c('#67001f','#b2182b','#d6604d',
                           '#f4a582','#fddbc7', '#f7f7f7',
                           '#d1e5f0','#92c5de','#4393c3',
                           '#2166ac', '#053061'))(50))
## draw heat map
pheatmap(data_norm[[1]], color=cols, border_color = NA, cutree_cols = 6)
```

```{r, results='hide', warning=F, message=F}
## draw heat map
pheatmap(data_norm[[2]], color=cols, border_color = NA, cutree_cols = 2)
```

```{r, results='hide', warning=F, message=F}
## draw heat map
pheatmap(data_norm[[3]], color=cols, border_color = NA, cutree_cols = 2)
```

```{r, results='hide', warning=F, message=F}
## draw heat map
pheatmap(data_norm[[4]], color=cols, border_color = NA, cutree_cols = 2)
```

# Session info

```{r}
sessionInfo()
```